# Reproducible Research: Peer Assessment 1


## Loading and preprocessing the data

```{r}
# Load Data if file exists else download and unzip data file 

if(!file.exists("data/activity.csv"))
{
  if (file.exists("activity.zip"))
  {
    unzip("activity.zip", exdir = "data")
  }
  else
  {
    print("Error : activity.zip file not found")
    cat("Download file [Y/N] ")
    response<-scan(n=1,what=character())
    if(response =="Y" || response=="y")
    {
     download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",destfile="activity.zip",method="curl")
     unzip("activity.zip", exdir = "data")
    }
    else
     { print("Script terminated")}
  } 
}
```
###Load Activity Data ###
  
  - With missiing values to adata
  
  - Tidy data values to atidydata
  
```{r}

# Load activity data into adata

adata<-read.csv("./data/activity.csv")
atidydata<-read.csv("./data/activity.csv",na.strings=c("NA","NULL"))
atidydata$ts<-as.POSIXct(as.double(atidydata$interval),origin=atidydata$date)

```

###Column Attributes###

```{r}
# Print variable names
names(adata)
```

###Summary###
```{r}
# Display summary
summary(adata)
```

### Process/Transform ###

```{r}
#ensure data types

class(adata$steps)
class(adata$date)
class(adata$interval)

```

## What is mean total number of steps taken per day?

### Mean total number of steps taken per slot and per day###
1. Make a histogram of the total number of steps taken each day
```{r}
#Calculate mean
library(plyr)
adatadaywise<-ddply(atidydata,~date,summarise,total=sum(steps),mean=mean(steps),median=median(steps),max=max(steps),lowest=min(steps))

#draw histrogram of steps per slot and per day side by side  
#par(mfrow=c(1,2))
hist(adata$steps,xlab="Total steps per slot",main="Histrogram of Total number of steps taken per slot")
hist(adatadaywise$total,xlab="Total steps per day",main="Histogram of Total Number of Steps taken each day")
```

###Mean and Median###
2. Calculate and report the mean and median total number of steps taken per day

```{r}
#display total and mean steps on each day of slots
###Total Steps,Mean(slots/day), Max and Lowest###
adatadaywise[,c("date","total","mean","median","max","lowest")]
adaysummary<-ddply(adatadaywise,~date,summarise,meanSteps=mean(total,na.rm=T),medianSteps=median(total,na.rm=T))
print(adaysummary)
```

```{r echo=FALSE}
cat("Mean:")
```
```{r echo=TRUE}
mean(adatadaywise$total,na.rm=T)
mean(adaysummary$meanSteps,na.rm=T)
```

```{r echo=FALSE}
cat("Median:")
```
```{r echo=TRUE}
median(adatadaywise$total,na.rm=T)
median(adaysummary$medianSteps,na.rm=T)

```

```{r echo=FALSE}
cat("Max:")
```
```{r echo=TRUE}
max(adatadaywise$total,na.rm=T)
```

## What is the average daily activity pattern?

1. Make a time series plot
   
```{r}

atidydata$ts<-as.POSIXct(as.double(atidydata$interval),origin=atidydata$date)
masteps<-mean(atidydata$steps,na.rm=T)
atidydata$mean <- masteps
mastepss<-mean(adaysummary$meanSteps,na.rm=T)
library(lattice)
xyplot(meanSteps~date,data=adaysummary,type=c("l","smooth"),panel=function(x,y,...){
     panel.xyplot(x,y,...)
     panel.abline(h=mastepss,col="red",lty="dotted")},xlab="interval",ylab="Avg Steps")
```

###Daily###

```{r}
xyplot(adatadaywise$mean~adatadaywise$date,type=c("l","smooth"),panel=function(x,y,...){
     panel.xyplot(x,y,...)
     panel.abline(h=mean(y,na.rm=T),col="red",lty="dotted")
 panel.abline(h=masteps,col="black",lty="dotted")},na.rm=T)

xyplot(steps~ts|date,data=atidydata,type="l",panel=function(x,y,...){
     panel.xyplot(x,y,...)
     panel.abline(h=mean(y,na.rm=T),col="red",lty="dotted")
 panel.abline(h=masteps,col="black",lty="dotted")},na.rm=T)
```

###Two months###

```{r}
xyplot(steps~ts,data=atidydata,type="l",panel=function(x,y,...){
     panel.xyplot(x,y,...)
     panel.abline(h=mean(y,na.rm=T),col="red",lty="dotted")
 panel.abline(h=masteps,col="black",lty="dotted")},na.rm=T)
```

###Subset - 29th of October 2012###

```{r}
xyplot(steps~ts|date,data=subset(atidydata,date == "2012-10-29"),type="l",panel=function(x,y,...){
     panel.xyplot(x,y,...)
     panel.abline(h=mean(y,na.rm=T),col="red",lty="dotted")
     panel.abline(h=masteps,col="black",lty="dotted")})
```

###Using qplot (ggplot)###

```{r}

library(ggplot2)
qplot(ts,steps,data=atidydata,geom=c("line","smooth"))

```

###One day in a month(2012-10-29)###

```{r}

qplot(ts,steps,data=subset(atidydata,date=="2012-10-29"),geom=c("line","smooth"))

```

2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r}

#print Max Steps, Date, Interval
atidydata[which.max(atidydata$steps),]
adata[which.max(adata$steps),]

```

## Imputing missing values

Calculate and report the total number of missing values in the dataset

```{r}
atidydataNA<- subset(atidydata,is.na(steps)|is.na(date)|is.na(interval))
cat("# no of rows with missing values ")
nrow(atidydataNA)
```

2. Device a strategy for filling in all of the missing values: Filled with mean value
   
   
```{r}
library(sqldf)
#update all the rows with 
atidydataWithoutNA<-sqldf(c("update atidydata set steps=mean where steps is null","select * from atidydata"))
#verify 
atidydataNA<- subset(atidydataWithoutNA,is.na(steps)|is.na(date)|is.na(interval))
nrow(atidydataNA)
```
3. New Data set has been created (atidydataWithoutNA) in previous step

4. Histogram 

```{r}

adataDayWiseWNA <- ddply(atidydataWithoutNA,~date,summarise,total=sum(steps),mean=mean(steps),median=median(steps),max=max(steps),lowest=min(steps))

#plot histogram
hist(adataDayWiseWNA$total)
#print values 
print(adataDayWiseWNA)

```

```{r echo=FALSE}
cat("Mean:")
```
```{r echo=TRUE}
mean(adataDayWiseWNA$total,na.rm=T)
```

```{r echo=FALSE}
cat("Median:")
```
```{r echo=TRUE}
median(adataDayWiseWNA$total,na.rm=T)
```

## Are there differences in activity patterns between weekdays and weekends?
What is the impact of imputing missing data on the estimates of the total daily number of steps?

The values are slightly different from the estimated value.

```{r}
finalDS<-data.frame(atidydataWithoutNA,ifelse(weekdays(as.Date(atidydataWithoutNA$date)) %in% c("Sunday","Saturday"),"Weekend","Weekday"))
#Appropriate field name
colnames(finalDS)[6]<-"Day"
#head
head(finalDS$Day)
#Enusure type is factor
class(finalDS$Day)
adatadaywise2<-ddply(finalDS,~date,summarise,total=sum(steps),mean=mean(steps),median=median(steps),max=max(steps),lowest=min(steps))
adaysummary2<-ddply(adatadaywise2,~date,summarise,meanSteps=mean(total),medianSteps=median(total))
#plot
meanstepss<-mean(adaysummary2$meanSteps,na.rm=T)
xyplot(steps~interval|Day,data=finalDS,type="l",layout=c(1,2))
xyplot(meanSteps~date,data=adaysummary2,type=c("l","smooth"),xlab="day",layout=c(1,2),panel=function(x,y,...){
     panel.xyplot(x,y,...)
 panel.abline(h=meanstepss,col="black",lty="dotted")})

xyplot(adatadaywise2$mean~adatadaywise2$date,type=c("l","smooth"),panel=function(x,y,...){
     panel.xyplot(x,y,...)
     panel.abline(h=mean(y,na.rm=T),col="red",lty="dotted")
 panel.abline(h=masteps,col="black",lty="dotted")},na.rm=T)
```
